---
title: "Class 05: Data Visualization with ggplot2"
author: "Kira Jung (PID: A16026398)"
format: html
---

### Question 1
> For which phases is data visualization important in our scientific workflows?

Answer: **All of the above** (Communication of results, EDA, detection of outliers).

### Question 2
> True or False? The ggplot2 package comes already installed with R?

Answer: **FALSE**.

### Question 3
> Which plot types are typically NOT used to compare distributions of numeric variables?

Answer: **Network graphs**.

### Question 4
> Which statement about data visualization with ggplot2 is incorrect?

Answer: "ggplot2 is the only way to create plots in R"

## Creating Plots (cars example)
First load the package and specify the dataset to be used as an argument.
> library (ggplot2), ggplot(cars), View(cars)

Create a basic plot

```{r}
plot(cars)
```

We can also specify aes and geom layers

```{r}
ggplot(data=cars) + aes(x=speed, y=dist) + geom_point()
```

We can add a line geom with geom_line()

```{r}
p <- ggplot(data=cars) + aes(x=speed, y=dist) + geom_point()

p + geom_line()

p + geom_smooth(method="lm")

```

## Read in our drug expression data
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
View(genes)

