---
title: "Class 6 Lab: R Functions"
author: "Kira Jung"
date: "April 25, 2023"
format: html
---

Introducing **R functions** and how to write our own R functions.

## Question 1

> Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

Writing a working code that solves a simple problem:

```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

mean(student1)
```
```{r}
# Which element of the vector is the lowest?
which.min(student1)

```

Now exclude the lowest score from the mean() calculation.

```{r}
# This returns everything but the eighth element of the vector.
student1[-8]
```

Now use the answer from which.min() to return all other elements of the vector.

```{r}
# This is our first working snippet
mean(student1[-which.min(student1)])
```

Now for the other example students, we can replace all NA values with zero

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x <- student2

x[is.na(x)] <- 0
x
mean(x)
```

Now dropping the lowest score for student two.

```{r}
mean(x[-which.min(x)])
```

Replacing NA with zero for student three.

```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

### Making the function
- A name, grade()
- Input arguments, a vector of student scores
- The body i.e. the working snippet of code

**Select 'Code > Extract Function'**

```{r}
#' Calculate the average score for a vector of student scores dropping the lowest score.
#' Missing values will be treated as zero.
#' 
#' @param x A numeric vector of homework scores
#'
#' @return Average score
#' @export
#'
#' @examples
#' student = c(100, NA, 90, 97)
#' grade(student)
#' 
grade <- function(x) {
  # Treat NA or missing values as zero to get an accurate score.
  
  x[is.na(x)] <- 0
  
  # Exclude lowest score from mean.
  
  mean(x[-which.min(x)])
}
```

```{r}
grade(student1)
grade(student2)
grade(student3)
```

Now using the function on our real class data from the CSV format file: "https://tinyurl.com/gradeinput”

```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)
```

```{r}
apply(gradebook, 1, grade)
```

## Question 2
> Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

Run the apply() function and save the results.

```{r}
results <- apply(gradebook, 1, grade)
sort(results, decreasing = TRUE)
```
```{r}
which.max(results)
```
### The highest scoring student overall is **student 18**.

## Question 3
> From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
gradebook
```

```{r}
ave.scores <- apply(gradebook, 2, mean, na.rm=TRUE)
ave.scores
which.min(ave.scores)
```

```{r}
med.scores <- apply(gradebook, 2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```

```{r}
boxplot(gradebook)
```

### Based on the median and boxplot, **homework 2** is the lowest scoring homework overall.

## Question 4
> From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

Are the final results (i.e. average score for each student) correlated with the results (i.e. scores) for individual homeworks..

```{r}
masked.gradebook <- gradebook
masked.gradebook[is.na(masked.gradebook)] <- 0
masked.gradebook
```

Then looking at correlation applied across the entire gradebook:

```{r}
apply(masked.gradebook, 2, cor, x=results) 
```

### Based on the results, **homework 5** was the most predictive of overall score based on its highest correlation.

## Question 5
> Make sure you save your Quarto document and can click the “Render” (or Rmark down”Knit”) button to generate a PDF format report without errors. Finally, submit your PDF to gradescope. [1pt