---
title: "Class 13 Mini Project"
author: "Kira Jung"
date: "2023-05-30"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Differential Expression Analysis

First loaded DESeq2 and then the data files.

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```

Deleting the first column

```{r}
countData <- as.matrix(countData[, -1])
```

Filtering count data where there is 0 read count across all samples

```{r}
countData = countData[rowSums(countData) > 1, ]
head(countData)
```

```{r}
library(DESeq2)
citation("DESeq2")
```


Setting up the DESeqDataSet object and running DESeq pipeline. First reading the metadata file.

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,colData=colData,design=~condition)
```


```{r}
dds <- DESeqDataSetFromMatrix(countData=countData,colData=colData,design=~condition)

dds
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)

res
```

```{r}
summary(res)
```


```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

Using the summary function on results to see how many genes are up or down regulated

```{r}
summary(res)
```

28% of the genes are up regulated and 29% of the genes are down regulated at the default 0.1 p-value cutoff.

Now making the volcano plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

Improving the plot by adding color and axis labels.

```{r}
# Making a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Genes with absolute fold change above 2
# colored red
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Genes with adjusted p-value less than 0.01
# and absolute fold change more than 2 color
# blue
inds <- (res$padj<0.01) & (abs(res$log2FoldChange) > 2)
mycols[ inds ] <- "blue"
plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

Adding gene annotation

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db, keys=row.names(res), keytype="ENSEMBL", column="SYMBOL", multiVals="first")

res$entrez = mapIds(org.Hs.eg.db, keys=row.names(res), keytype="ENSEMBL", column="ENTREZID", multiVals="first")

res$name = mapIds(org.Hs.eg.db, keys=row.names(res), keytype="ENSEMBL", column="GENENAME", multiVals="first")

head(res, 10)
```

Reordering the results by adjusted p-value
Save to CSV file

```{r}
res = res[order(res$padj),]
ord <- order( res$padj )
head(res[ord,])
write.csv((res[ord,]), file = "deseq_results.csv")
```

# Section 2. Pathway Analysis

After installing bioconductor packages:

```{r}
library(pathview)
```

```{r}
library(gage)
```

```{r}
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

head(kegg.sets.hs, 3)
```

Running the gage pathway analysis:

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
head(keggres$less)
```

Using the pathview funtion to make a pathway plot:

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

Changing the display to generate PDF graph.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

Focusing on the top 5 upregulated pathways:

```{r}
keggrespathways <- rownames(keggres$greater)[1:5]

keggresids = substr(keggrespathways, start=1, stop=8)

keggresids
```

Drawing plots for all top 5 upregulated pathways:

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

Focusing on top 5 downregulated pathways:

```{r}
keggrespathways_downregulated <- rownames(keggres$less)[1:5]

keggresids_downregulated = substr(keggrespathways_downregulated, start=1, stop=8)
keggresids_downregulated
```

Drawing plots for all top 5 downregulated pathways

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids_downregulated, species="hsa")
```

# Section 3. Gene Ontology

```{r}
data(go.sets.hs)

data(go.subs.hs)

# Focusing on Biological Process subset
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

# Section 4. Reactome Analysis

Outputting the list of significant genes at the 0.05 level as a plain text file.

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]

print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

The pathway with the most significant entities p-value is cell cycle, mitotic. The previous KEGG results shows the M phase as being the pathway with the most significant p-value. The difference in the results could be due to the fact that KEGG uses more broad terms while Reactome has similar terms as multiple entries. 

```{r}
sessionInfo()
```

